From physical principles to ???

### 2016-03-09 2015CompleteDerivation/completeDerivation.tex
Reviewed state definitions

\section{States}

@@

Before we are able to define how states should be captured mathematically, we need to better specify what we mean by physically distinguishable. Therefore we need to characterize what we mean by measurement, particularly because it has become such an overloaded term.

Consider the motion of a cannonball under inertial and gravitational forces. Light will be scattering off of it; as it lands the ground will be deformed and its temperature will rise. Those external physical processes, which happen no matter what we do, can be used to distinguish the motion of the cannonball as their outcomes are correlated. Therefore we can learn the position by looking at the reflected light, and learn the final kinetic momentum by looking at the deformation of the ground. In principle, any external process that has a correlation with the states under study can be used to perform a measurement, and any measurement is based on such a process. That is: for us a measurement is simply a physical process that we can use to distinguish states. There is no particular role played by the observer, except maybe create the device and forcing the system under study to interact with it.

This external process may be quite complicated: when a particle enters a calorimeter, a shower of particles is produced, photons are captured and are directed to photomultipliers, a current is read out, the current is then digitized, and so on. Some processes interfere with the system, they affect its dynamics, and other are destructive, the system no longer can be described by the original set of states. A tracking chamber is an example of the first (the magnetic field curves the motion of a charged particle); burning a substance to determine its caloric content is an example of the latter. In these cases, special care is needed to make sure to link the outcomes with the original state, unaffected state. But in the end, the conceptual description remains the same.

%TODO: It's not about precision! Discrete variables are known to infinite precision! Is about information.

We have a physical process that provides an outcome. This process has a set of potential outcomes. To each potential outcome we can associate a set of states. For example, if the electron follows a certain path, then its state is among the ones that has spin $+1/2$. Since all measurements are finite precision, each outcome can be labeled by a finite set of digits, so the set of outcomes is countable. Note that different outcomes for the same process can overlap. For example, $4.12 \pm 0.05$ cm and $4.13 \pm 0.05$ cm are both legitimate possible outputs of the same measurement device. But since all states must be distinguishable, given two arbitrary states there must be a process precise enough to tell them apart. That is: the potential outcomes associated with the two states do not overlap. For example, $4.12 \pm 0.0005$ cm and $4.13 \pm 0.0005$ cm do not overlap anymore.

%TODO If we start "combining processes" then is it really _a_ process?

We can also combines two outcome of two different processes. For example, if one tells us spin is $+1/2$ and another that position is $4.13 \pm 0.05$ cm, then we can say the state has both. That is, the state is inside both of the sets of states associated with the two outcomes: the intersection of two sets associated with an outcome is also a set associated with an outcome. We can make similar arguments about the union as we can always reduce the precision by combining different outcome of the same process.

@@

With our state space defined, deterministic and reversible evolution corresponds to a bijective map between initial and final states.

TODO: det/rev map from manifold to manifold. Must be physically distinguishable. Can't be the whole class of functions: refining domain does not refine codomain. (mathematically, cardinality greater than continuoum, i.e. not second countable) We need that arbitrarily refining domain refines the codomain. Equivalent to continuous function(?). OR Measurement/process outcomes from initial states corresponds to outcomes on final states. That is: we map open sets to open sets.

\begin{defn}\label{detrevmap}
A deterministic and reversible evolution is a self-homeomorphism on the state space $\mathcal{S}$.
\end{defn}

\begin{justification}
The system is deterministic: there exists a map $f:\mathcal{S} \rightarrow \mathcal{S}$ that returns the final state given the initial state. The system is reversible: there exists a map $g:\mathcal{S} \rightarrow \mathcal{S}$ that returns the initial state given the final state. $g \circ f = id$ as mapping forward and then backward must return the original element.

TODO: show it maps open sets to open sets.
\end{justification}

>>>>

\section{States and state space}

@@

We now proceed to characterize states and their being physically distinguishable in more precise terms so that we can capture their description mathematically. What we'll see is that the set of all physical processes that can be used to gain information about the system, that can be used to perform a measurement, induce a topology on the set of states that is Hausdorff and second countable. That is: physical distinguishability  is mathematically captured by topological distinguishability. Deterministic and reversible evolution will then preserve the topology, and they will be mathematically captured by self-homeomorphisms.

We'll focus on state spaces that can be described by a set of independent state variables, either discrete or continuous, and see what can be said in general on the evolution of state variables.

\subsection{States and topology}

As the term measurement has become a particularly loaded, let's first characterize what we mean by physical distinguishability in our context.

Consider the motion of a cannonball under inertial and gravitational forces. Light will scatter off of it; as it lands the ground will be deformed and the impact will make the temperature rise. Those external physical processes, which happen no matter what we do, can be used to distinguish the motion of the cannonball as their outcomes are correlated.\footnote{Note that by requiring a correlation we are implying that measurements are repeatable. That is, for similarly prepared states we get similar outcomes.} Therefore we can learn the position by looking at the reflected light, and learn the final kinetic energy by looking at the deformation of the ground. In principle, any external process that has a correlation with the states under study can be used to perform a measurement, and any measurement is based on such a process. That is: for us a measurement is simply a physical process that we can use to distinguish states. There is no particular role played by the observer, except maybe setting up such a process and forcing the system under study to interact with it.

This external process may be quite complicated: when a particle enters a calorimeter, a shower of particles is produced, photons are captured and are directed to photomultipliers, a current is read out, the current is then digitized, and so on. Some processes interfere with the system, they affect its dynamics, and other are destructive, the system no longer can be described by the original set of states. A tracking chamber is an example of the first (the magnetic field curves the motion of a charged particle); burning a substance to determine its caloric content is an example of the latter. In these cases, special care is needed to make sure to link the outcomes with the original state. But in the end, the conceptual description remains the same: each possible outcome of the process will be associated with a set of states consistent with that outcome (e.g. if the cannonball deformed the ground by this amount than the final state its kinetic energy at impact was within this range).

Each physical process, then, can be thought as having a set of potential outcomes. To each potential outcome we can associate a set of states that are consistent with it. For example, if the electron follows a certain path, then its state is among the ones that has spin $+1/2$. However precise our measurements are, each outcome is expressed by a finite set of digits, therefore the set of outcomes is countable. Note that different outcomes for the same process can overlap. For example, $4.12 \pm 0.05$ cm and $4.13 \pm 0.05$ cm are both legitimate possible outputs of the same measurement device. But since all states must be distinguishable, given two arbitrary states there must be a process precise enough to tell them apart. That is: the potential outcomes associated with the two states do not overlap. For example, $4.12 \pm 0.0005$ cm and $4.13 \pm 0.0005$ cm do not overlap anymore.

We can also conceptually combine two different processes into a single one. That is: having a way to measure quantity $x$ and a way to measure quantity $y$ gives us a way to measure the combination $(x,y)$. Formally, the outcomes of the combined process will be the intersections of each pair of outcomes of the original processes.

We can also coarsen a single process. That is: having a way to measure $(x,y)$ gives us a way to measure $x$ alone (or any $f(x,y)$). Formally, the outcomes of the coarsened process are given by performing the union of some outcomes of the original process.

@@

With our state space defined, deterministic and reversible evolution corresponds to a bijective map between initial and final states. Moreover, a physical process that can distinguish initial states is also a physical process that can distinguish final states. That is: the topology is mapped and preserved by the deterministic and reversible evolution (i.e. initial and final states are equally distinguishable). The evolution is then a self-homeomorphism on the state space.

\begin{defn}\label{detrevmap}
A deterministic and reversible evolution map is a self-homeomorphism on the state space $f:\mathcal{S} \rightarrow \mathcal{S}$.
\end{defn}

\begin{justification}
We claim that $f$ exists. The system is deterministic: given an initial state $\mathcal{s} \in \mathcal{S}$ there exists a well defined final state $f(\mathcal{s})=\hat{\mathcal{s}} \in \mathcal{S}$.

We claim that $f$ is a bijection. The system is reversible: there exists a map $g:\mathcal{S} \rightarrow \mathcal{S}$ that returns the initial state given the final state. $g \circ f = f \circ g = id$ as mapping forward and then backward or backwards and then forward must return the original element. $f$ admits $g$ as an inverse. $f$ is a bijection

We claim $f$ is an open map. Let $U \subseteq \mathcal{S}$ represent an outcome of a process. $U$ is an open set in the state space topology by definition. Given the deterministic and reversible evolution $f$, an outcome on the initial states can be used to distinguish final states. The same outcome associated with $U$ as initial states will also be associated with $f(U)$ as final states. $f(U)$ is associated with an outcome. $f(U)$ is an open set in the state space topology by definition.
\end{justification}

%TODO: do continuous function form a manifold Hausdorf and second countable?

\subsection{Labeling states}
